---
title: "gamma_approx_to_nb"
author: "Mangiola Stefano"
date: "15/06/2018"
output:
  github_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: 3
    code_folding: hide
linkcolor: magenta
urlcolor: magenta
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, results='hide'}
library(foreach)
library(tidyverse)

```

## Comparison of overdispersion between gamma, negative binomial, with poisson

Both negative binomial and gamma can limit to a poisson

```{r limit_to_poisson}

mm = c(seq(2, 10, 1), seq(10, 5000, 10))
my_df = 
	
	# Gamma
	foreach(m = mm, .combine = bind_rows) %do% {
	s=1

	x = rgamma(10000,shape = m * s, rate = s)
	
	c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	} %>% 
	rename("gamma" = `spread.10%`) %>% 
	
	# Negative binomial
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=1000000
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		} %>% 
			rename("nb" = `spread.10%`),
	by="mean"
	) %>%
		
	# Poisson
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			
			x = rpois(10000, lambda = m)
			
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		} %>% rename("pois" = `spread.10%`),
		by="mean"
	) %>%
		gather(label, spread, -mean) 

```


```{r, echo=FALSE}
my_df %>% ggplot(aes(x=mean, y=spread, color=label)) + geom_line() + theme_bw() 
```


However gamma and negative binomial introduce overdispersion "generalising" the relation between expected value and variance of the poisson in a different way

This is the gamma with increasing overdispersion, from poisson.

```{r gamma}

mm = c(seq(2, 10, 1), seq(10, 5000, 10))
my_df = 
	
	# Gamma
	foreach(m = mm, .combine = bind_rows) %do% {
		s=1
		x = rgamma(10000,shape = m * s, rate = s)
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	} %>% 
	rename("1" = `spread.10%`) %>% 
		
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=0.8
			x = rgamma(10000,shape = m * s, rate = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("0.8" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=0.6
			x = rgamma(10000,shape = m * s, rate = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("0.6" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=0.4
			x = rgamma(10000,shape = m * s, rate = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("0.4" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=0.2
			x = rgamma(10000,shape = m * s, rate = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("0.2" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=0.1
			x = rgamma(10000,shape = m * s, rate = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("0.1" = `spread.10%`),
		by="mean"
	) %>%
		gather(rate_param_R_gamma, spread, -mean) 

```


```{r, echo=FALSE}
my_df %>% ggplot(aes(x=mean, y=spread, color=rate_param_R_gamma)) + geom_line() + theme_bw() 
```


This is the negative binomial with increasing overdispersion, from poisson.

```{r negative_binomial}

mm = c(seq(2, 10, 1), seq(10, 5000, 10))
my_df = 
	
	# Negative binomial
	foreach(m = mm, .combine = bind_rows) %do% {
			s=1000000
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		} %>% 
	rename("1000000" = `spread.10%`) %>% 
		
	# Negative binomial
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=100000
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("100000" = `spread.10%`),
		by="mean"
	) %>%
	
	# Negative binomial
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=10000
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		} %>% rename("10000" = `spread.10%`),
		by="mean"
	) %>%
	
	# Negative binomial
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=1000
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("1000" = `spread.10%`),
		by="mean"
	) %>%
	
	# Negative binomial
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=100
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("100" = `spread.10%`),
		by="mean"
	) %>%
	
	# Negative binomial
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=10
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		} %>% rename("10" = `spread.10%`),
		by="mean"
	) %>%
		gather(size_param_R_nb, spread, -mean) 

```


```{r, echo=FALSE}
my_df %>% ggplot(aes(x=mean, y=spread, color=size_param_R_nb)) + geom_line() + theme_bw() 
```

If we superimpone gamma and negative binomial

```{r gamma and negative_binomial}

mm = c(seq(2, 10, 1), seq(10, 5000, 10))
my_df = 
	
	# Negative binomial
	foreach(m = mm, .combine = bind_rows) %do% {
			s=3000
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		} %>% rename("negative binomial" = `spread.10%`) %>%
	
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=0.4
			x = rgamma(10000,shape = m * s, rate = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("gamma" = `spread.10%`),
		by="mean"
	) %>% 
	gather(Label, spread, -mean) 

```


```{r, echo=FALSE}
my_df %>% ggplot(aes(x=mean, y=spread, color=Label)) + geom_line() + theme_bw() 
```

Now let's try to reparametrise the gamma to have constant SD and apply a linear dependence between SD and mean

```{r gamma reparametrised}

mm = c(seq(2, 10, 1), seq(10, 5000, 10))
my_df = 
	
	# Gamma
	foreach(m = mm, .combine = bind_rows) %do% {
		s=1
		s = s + 0.3 * m
		
		v = s^2;
		shape = m^2 / v;
		rate = m / v;
		
		x = rgamma(10000,shape = shape, rate = rate)
			
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	} %>% 
	rename("0.3" = `spread.10%`) %>% 
		
	# Gamma
	left_join(
	foreach(m = mm, .combine = bind_rows) %do% {
		s=1
		s = s + 0.25 * m
		
		v = s^2;
		shape = m^2 / v;
		rate = m / v;
		
		x = rgamma(10000,shape = shape, rate = rate)
			
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	} %>% rename("0.25" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
	foreach(m = mm, .combine = bind_rows) %do% {
		s=1
		s = s + 0.20 * m
		
		v = s^2;
		shape = m^2 / v;
		rate = m / v;
		
		x = rgamma(10000,shape = shape, rate = rate)
			
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	}  %>% rename("0.20" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
	foreach(m = mm, .combine = bind_rows) %do% {
		s=1
		s = s + 0.15 * m
		
		v = s^2;
		shape = m^2 / v;
		rate = m / v;
		
		x = rgamma(10000,shape = shape, rate = rate)
			
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	}  %>% rename("0.15" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
	foreach(m = mm, .combine = bind_rows) %do% {
		s=1
		s = s + 0.10 * m
		
		v = s^2;
		shape = m^2 / v;
		rate = m / v;
		
		x = rgamma(10000,shape = shape, rate = rate)
			
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	} %>% rename("0.10" = `spread.10%`),
		by="mean"
	) %>%
	
	# Gamma
	left_join(
	foreach(m = mm, .combine = bind_rows) %do% {
		s=1
		s = s + 0.05 * m
		
		v = s^2;
		shape = m^2 / v;
		rate = m / v;
		
		x = rgamma(10000,shape = shape, rate = rate)
			
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	}  %>% rename("0.05" = `spread.10%`),
		by="mean"
	) %>%
		gather(multiplic_factor, spread, -mean) 

```


```{r, echo=FALSE}
my_df %>% ggplot(aes(x=mean, y=spread, color=multiplic_factor)) + geom_line() + theme_bw() 
```

If we superimpone gamma and negative binomial

```{r gemme reparam, gamma and negative_binomial}

mm = c(seq(2, 10, 1), seq(10, 5000, 10))
my_df = 
	
	# Negative binomial
	foreach(m = mm, .combine = bind_rows) %do% {
			s=3000
			x = rnbinom(10000, mu = m, size = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		} %>% rename("negative binomial" = `spread.10%`) %>%
	
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
		s=10
		s = (s + 0.02 * m) 
		
		v = s^2;
		shape = m^2 / v;
		rate = m / v;
		
		x = rgamma(10000,shape = shape, rate = rate)
			
		c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
	}  %>% rename("gamma_reparam" = `spread.10%`),
		by="mean"
	) %>% 
	
	# Gamma
	left_join(
		foreach(m = mm, .combine = bind_rows) %do% {
			s=0.4
			x = rgamma(10000,shape = m * s, rate = s)
			c(mean = m, spread = abs(quantile(x, 0.1) - quantile(x, 0.9)))
		}  %>% rename("gamma" = `spread.10%`),
		by="mean"
	) %>% 
	gather(Label, spread, -mean) 

```


```{r, echo=FALSE}
my_df %>% ggplot(aes(x=mean, y=spread, color=Label)) + geom_line() + theme_bw() 
```
